---
title: "Create units for monarchs"
author: "Christen Bossu, modified by W. Chris Funk"
date: "4/22/2025, modified by W. Chris Funk on 09/12/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Making units file

And now we can start processing that stuff.  Since the file names
don't all follow the same conventions, we have to do slightly different
things with the `.fq.gz` ones versus the `.fastq.gz` ones.  We throw in
a little `case_when()` for those situations.

Note that here- the naming convention for the three lanes were the same, so I simplified this code chunk. If you need the case_when() function I can send you an updated code! I got this file by using the ls -l command to the location of all the fastq files that I will be using- I had 3 folders


```
ls -l --color=never data/fastqs/*/*gz > monarch.fastq_listing.txt
```

I had to make separate lists because sample ID names have different numbers of "_" in them. Total pain in the ass, so in future, need to make sure that sample ID names have the SAME number of "_" in their names.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)

files1underscores <- read_table("fastq_listing.underscores-1.txt", col_names = FALSE) %>%
  select(X9, X5) %>%
  mutate(kb = X5/1000) %>%
  rename(fq = X9) %>%
  select(fq, kb) %>%
  mutate(base = basename(fq)) %>% 
  separate(base,into=c("N1","read"),sep="[_.]") %>% mutate(sample_id = paste(N1, sep = "_"))

files2underscores <- read_table("fastq_listing.underscores-2.txt", col_names = FALSE) %>%
  select(X9, X5) %>%
  mutate(kb = X5/1000) %>%
  rename(fq = X9) %>%
  select(fq, kb) %>%
  mutate(base = basename(fq)) %>% 
  separate(base,into=c("N1","N2","read"),sep="[_.]") %>% mutate(sample_id = paste(N1, N2, sep = "_"))

files3underscores <- read_table("fastq_listing.underscores-3.txt", col_names = FALSE) %>%
  select(X9, X5) %>%
  mutate(kb = X5/1000) %>%
  rename(fq = X9) %>%
  select(fq, kb) %>%
  mutate(base = basename(fq)) %>% 
  separate(base,into=c("N1","N2","N3","read"),sep="[_.]") %>% mutate(sample_id = paste(N1, N2, N3, sep = "_"))

files4underscores <- read_table("fastq_listing.underscores-4.txt", col_names = FALSE) %>%
  select(X9, X5) %>%
  mutate(kb = X5/1000) %>%
  rename(fq = X9) %>%
  select(fq, kb) %>%
  mutate(base = basename(fq)) %>% 
  separate(base,into=c("N1","N2","N3","N4","read"),sep="[_.]") %>% mutate(sample_id = paste(N1, N2, N3, N4, sep = "_"))
  
files5underscores <- read_table("fastq_listing.underscores-5.txt", col_names = FALSE) %>%
  select(X9, X5) %>%
  mutate(kb = X5/1000) %>%
  rename(fq = X9) %>%
  select(fq, kb) %>%
  mutate(base = basename(fq)) %>% 
  separate(base,into=c("N1","N2","N3","N4","N5","read"),sep="[_.]") %>% mutate(sample_id = paste(N1, N2, N3, N4, N5, sep = "_"))
  
files6underscores <- read_table("fastq_listing.underscores-6.txt", col_names = FALSE) %>%
  select(X9, X5) %>%
  mutate(kb = X5/1000) %>%
  rename(fq = X9) %>%
  select(fq, kb) %>%
  mutate(base = basename(fq)) %>% 
  separate(base,into=c("N1","N2","N3","N4","N5","N6","read"),sep="[_.]") %>% mutate(sample_id = paste(N1, N2, N3, N4, N5, N6, sep = "_"))
  
files1underscores <- files1underscores %>% mutate(source = "oneunderscores")
files2underscores <- files2underscores %>% mutate(source = "twounderscores")
files3underscores <- files3underscores %>% mutate(source = "threeunderscores")
files4underscores <- files4underscores %>% mutate(source = "fourunderscores")
files5underscores <- files5underscores %>% mutate(source = "fiveunderscores")
files6underscores <- files6underscores %>% mutate(source = "sixunderscores")

files <- bind_rows(files1underscores, files2underscores, files3underscores, files4underscores, files5underscores, files6underscores)

files$sample_id
```

Now, because the naming of the files does not always have the lane, etc., and
we want to machine name (not really necessary, but we can get it easily from
the first line of each file), I make a file that has that info that we can join
onto the path.

```sh
for i in data/fastqs/*/*gz; do zcat $i | awk -v f=$i 'BEGIN {OFS="\t"} NR==1 {print f, $1; exit}'; done | awk '!/Undetermined/' > monarch.novo.seq-tags-per-path.tsv
```
The results of that have been put into:
```
example-configs/monarchs/prep/monarch.novo.seq-tags-per-path.tsv
```
And we can make it nice like this:
```{r}
seq_ids <- read_tsv("monarch.novo.seq-tags-per-path.tsv", col_names = c("fq", "id")) %>%
  separate(
    id, 
    into = c("x1", "x2", "flowcell", "lane"), 
    sep = ":", 
    extra = "drop"
  ) %>%
  dplyr::select(-x1, -x2) %>%
  mutate(platform = "ILLUMINA")


seq_ids %>% distinct(flowcell)
```

And we can now join those and pivot them to get fq1 fq2 kb1 and kb2 all on the same
line, and then assign snakemake sample numbers to them.
```{r}

library(dplyr)
library(stringr)
library(tidyr)
library(readr)
library(purrr)

# 1) Make sure reads are standardized to "1"/"2"
files <- files %>%
  mutate(read = recode(read, "R1" = "1", "R2" = "2")) %>%
  filter(read %in% c("1", "2"))

# 2) Join header-derived flowcell/lane
files_joined <- files %>%
  left_join(seq_ids, by = "fq")

# (Optional) quick sanity checks you can run once:
# -- any paths with no header match?
# anti_join(files, seq_ids, by = "fq") %>% count()
# -- any weird read labels left?
# files_joined %>% count(read)

# 3) Deduplicate within (sample_id, flowcell, lane, read)
#    keep subdirectory identity so intralib_reps doesn't collapse into libX
files_dedup <- files_joined %>%
  mutate(dir = dirname(fq)) %>%
  arrange(dir, lane, desc(kb)) %>%
  distinct(sample_id, flowcell, lane, read, dir, .keep_all = TRUE)

# 4) Wide pivot -> fq1, fq2, kb1, kb2 (now they’ll be scalars, not lists)
files_wide <- files_dedup %>%
  pivot_wider(
    id_cols    = c(sample_id, flowcell, lane, platform, dir), # <- dir separates replicates
    names_from = read,
    values_from = c(fq, kb),
    names_sep  = ""
  ) %>%
  arrange(sample_id, flowcell, lane, dir)

# 5) Stable sample codes (keeps your order-of-first-appearance behavior)
files_wide <- files_wide %>%
  mutate(
    sample = sprintf(
      "s%04d",
      as.integer(factor(sample_id, levels = unique(sample_id)))
    ),
    .before = sample_id
  )

# 6) Map flowcell -> library with a clean lookup (same mapping you had)
flowcell_to_library <- c(
  "HCVC5DSX7" = "lib1",
  "22HM5LLT4" = "lib2",
  "22HVG3LT4" = "lib3",
  "22M5VMLT4" = "lib4",
  "22T3TYLT4" = "lib5",
  "HLYTNAFXY" = "deroode1",
  "HLTC7AFXY" = "deroode2",
  "HM3NJAFXY" = "deroode3",
  "AAAK73JM5" = "deroode4",
  "AAAKJL3M5" = "deroode5",
  "AAAJVLMM5" = "deroode6"
)

files_wide <- files_wide %>%
  mutate(library = recode(flowcell, !!!flowcell_to_library, .default = "deroode7"))

```


So, now we can make our units file.  For barcodes I am just going to do the
sample_id + library + lane.  It just needs to be unique for those individuals.
```{r}

units <- files_wide %>%
  arrange(sample, library, lane, dir, fq1) %>%   # dir distinguishes intralib reps
  group_by(sample) %>%
  mutate(unit = row_number(), .after = sample) %>%
  ungroup() %>%
  mutate(barcode = stringr::str_c(sample_id, library, lane, sep = "-")) %>%
  select(sample, unit, library, flowcell, platform, lane,
         sample_id, barcode, fq1, fq2, kb1, kb2) %>%
  arrange(sample, unit)

write_tsv(units, "monarch.units.tsv")

```


## Making chromosomes and scaffold groups

We do this with R.  As always, it is important to look at the format
in `.test/chromosomes.tsv` and `.test/scaffold_groups.tsv` to know the format.
```{r}
#BLVI ragtag genome: GCA_013396875.1_ASM1339687v1_scaffolded_ragtag_genomic.fna.fai

# as I did before, we will let anything over 30 Mb be a "chromosome" and then
# we will shoot for scaffold groups < 50 Mb in total.
fai <- read_tsv(
  "~/Dropbox/BGP/WAVI/WAVI_genome/Vireo_gilvus_B10K_scaffolded.fasta.gz.fai", col_names = c("chrom", "len", "x1", "x2", "x3")) %>%
  dplyr::select(-starts_with("x")) %>%
  mutate(cumlen = cumsum(len))

# here are the lengths:
fai %>%
  mutate(x = 1:n()) %>%
  ggplot(aes(x=x, y = len)) + geom_col()
```
Proceeding:
```{r}
chromos <- fai %>%
  filter(len >= 4e6) %>%
  rename(num_bases = len) %>%
  dplyr::select(-cumlen)

write_tsv(chromos, file = "WAVI.chromosomes.ragtag.tsv")

loco_use <- fai %>%
  filter(len >= 100000) %>%
  rename(num_bases = len) %>%
  dplyr::select(-cumlen,-num_bases)

write_tsv(loco_use, file = "WAVI.chromosomes.100Kb.ragtag.tsv")

# now, get the scaff groups
scaffs <- fai %>%
  filter(len < 4e6)

bin_length <- 2e06

scaff_groups <- scaffs %>%
  mutate(
    cumul = cumsum(len),
    part = as.integer(cumul / bin_length) + 1L
  ) %>%
  mutate(
    id = sprintf("scaff_group_%03d", part),
    .before = chrom
  ) %>%
  dplyr::select(-part, -cumlen)

# let's just see the total lengths of those scaff_groups
# and also the number of scaffolds in each
scaff_groups %>%
  #read_delim("ROFI.scaffold_groups.tsv",delim="\t") %>% 
  group_by(id) %>%
  summarise(
    tot_bp = sum(len),
    num_scaff = n()
  ) %>% print(n=35)
```
Good, that is not too many scaff groups, and also not too many scaffolds
per any one group.
```{r}
write_tsv(scaff_groups, file = "WAVI.scaffold_groups.ragtag.tsv")
```


## Setting up the config.yaml file

As always, it is important to start from .test/config/config.yaml, because that
is the most up-to-date.  So I copied that to 
```
example-configs/YEWA-Jan-2023/config.yaml
```
And then I edited it as appropriate.

## Getting the scatters file

I set the scatters to "" in the config like this:
```yaml
scatter_intervals_file: ""
```
And then I committed all the stuff I just created here and pushed it up to the cluster.
Then, as suggested in the README, did this:
```sh
(base) [node21: mega-non-model-wgs-snakeflow]--% pwd
/home/eanderson/scratch/PROJECTS/YEWA-Jan-2023/mega-non-model-wgs-snakeflow
(base) [node21: mega-non-model-wgs-snakeflow]--% conda activate snakemake-7.7.0
(snakemake-7.7.0) [node21: mega-non-model-wgs-snakeflow]--% module load R
(snakemake-7.7.0) [node21: mega-non-model-wgs-snakeflow]--% snakemake  --cores 1 --use-conda results/scatter_config/scatters_5000000.tsv --configfile example-configs/WIFL-April-2025/config.yaml

# then copy the result to the config area:
(snakemake-7.7.0) [node21: mega-non-model-wgs-snakeflow]--% cp results/scatter_config/scatters_5000000.tsv example-configs/YEWA-Jan-2023/
```
After that I also updated the config file to point to that new file:
```yaml
scatter_intervals_file: "example-configs/YEWA-Jan-2023/scatters_5000000.tsv"
```

Now, when it is calling variants from the Genomics Data Bases, it will scatter
the job over sections of genome that are less that 5 Mb (hence the 5000000 in the
name).

I committed that and pushed it back up.

And now we are ready to do a dry run:
```sh
(snakemake-7.7.0) [node21: mega-non-model-wgs-snakeflow]--% snakemake -np --profile hpcc-profiles/slurm/sedna 
YEWA-Jan-2023/config.yaml
```


We can check that this makes sense with the numbers of things we have:
```{r}
nrow(units_all)
```

35 units.  That is the number of trimming and mapping and fastqc jobs we expect.

```{r}
n_distinct(units_all$sample)
```

35 distinct samples.  That is the number of dup-marking and other sorts of
jobs we expect.

```{r}
nrow(chromos)
n_distinct(scaff_groups$id)
```

1 chromosome and 21 scaffold groups.  So we expect a few jobs doing
1 or 21, or 25 things.

And, for making gvcf sections, we expect:
Number of individuals times (number of chroms + number of scaff groups) so:
22 * 35 = 770.

Finally, we expect the total number of GenomicsDB2VCF steps to be the number
of unique scatter groups (Note that we use the old scatters file here, so that
things stay consistent when I run this notebook.
```{r}
scats <- read_tsv("../scatters_5000000.tsv")
n_distinct(scats$id, scats$scatter_idx)
```

So, 274 of those jobs.  
